---
title: "Tidy Tuesday - Week 4: The Exploding Snowman Predicts Summer"
author: "Sk Abidur Rahman"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
format:
  html:
    toc: true
    theme: flatly
execute:
  warning: false
  message: false
---

## Overview

The new technique I learned this week is calculating a **Rolling Correlation** using the **`slider`** package.

* **What it does:** Instead of calculating one correlation for the entire 100-year history, I can calculate a correlation for every successive 15-year period (e.g., 1950-1964, then 1951-1965, etc.).
* **Why it's important:** This technique allows us to find a small window of time where the snowman's prediction might have been temporarily more accurate, which is necessary to answer one of the assignment questions.

---

## 🛠️ Setup & File Structure (2 Points)

This chunk loads the necessary toolkits and prepares the file system.

```{r}
# packages 
library(tidyverse) #core package for data manipulation (dplyr) and plotting (ggplot2)
library(here)      # define file paths consistently
library(slider)    # for the Rolling Correlation 

# path to the data file
f_data <- here("Tidy_Tuesday","Week_4", "data", "sechselaeuten.csv")

# Output folder
dir.create(here("Tidy_Tuesday","Week_4","output"))


# Read the dataset
snowman <- readr::read_csv(f_data, show_col_types = FALSE)

# Check
head(snowman)
```


#Main Analysis: Duration vs. Summer Temperature
```{r}
# Data Clean
boeoegg_data <- snowman %>%
  # Select the three columns: year, burn duration, and average temperature
  select(year, duration, avg_temp = tre200m0) %>%
  # Remove missing (NA) value
  filter(!is.na(duration) & !is.na(avg_temp))

# Calculate the Correlation (r) 
# Correlation measures the linear relationship (-1 is perfect negative, 0 is no relation, 1 is perfect positive)
correlation_r <- cor(boeoegg_data$duration, boeoegg_data$avg_temp)

# Scatter Plot
plot_main <- boeoegg_data %>%
  # X is burn duration, Y is average temperature
  ggplot(aes(x = duration, y = avg_temp)) +
  # Add the data points 
  geom_point(color = "#005a9c", alpha = 0.7, size = 3) +
  # Add a linear model 'lm' trend line to show the overall pattern
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 1) +
  #titles and labels
  labs(
    title = "Does the Exploding Snowman Predict Summer Temperature?",
    subtitle = paste("Correlation (r) =", round(correlation_r, 2), ". The myth seems unsupported by data."),
    x = "Boeoegg Burn Duration (Minutes)",
    y = "Average Summer Temperature (°C)"
  ) +
  theme_minimal(base_size = 14)

# Show the plot
plot_main

# Save 
ggsave(
  filename = here("output", "snowman_vs_temperature.png"),
  plot = plot_main,
  width = 8, height = 5, dpi = 320, create.dir = TRUE
)
```


#Successive Years with Higher Accuracy
```{r}
#rolling-correlation
# Define the rolling window size (years) 
window_size <- 15  # 15-year rolling window

# Calculate the rolling correlation between duration and avg_temp 
rolling_cor <- boeoegg_data %>%
  arrange(year) %>%        # ensure data is ordered by year
  mutate(
    rolling_r = slider::slide2_dbl(
      .x = duration,       # first series
      .y = avg_temp,       # second series
      .f = ~ cor(.x, .y),  # correlation 
      .before = window_size - 1,  # look back 14 years + current = 15
      .complete = TRUE              # only compute when window is full
    )
  ) %>%
  filter(!is.na(rolling_r))  # drop leading NAs where window isn't full

# Find the most negative or strongest inverse in 15-year period
most_accurate_period <- rolling_cor %>%
  slice_min(rolling_r, n = 1)

most_accurate_period
```


# Forecasting Ability with Other Climate Variables
```{r}
# Select burn time and all climate variables columns
climate_correlation_data <- snowman %>%
  # Select duration and the mean summer columns for Temp, Sunshine, and Rain
  select(duration, tre200m0, sre000m0, rre150m0) %>%
  # Keep only years where all data is available
  filter(complete.cases(.)) 

#Calculate the R-squared 
# R-squared is the percentage of variation explained by the linear model

# Model 1: Duration vs. Average Temperature (tre200m0)
r_sq_temp <- summary(lm(tre200m0 ~ duration, data = climate_correlation_data))$r.squared

# Model 2: Duration vs. Total Sunshine Duration (sre000m0)
r_sq_sun <- summary(lm(sre000m0 ~ duration, data = climate_correlation_data))$r.squared

# Model 3: Duration vs. Total Precipitation (rre150m0)
r_sq_rain <- summary(lm(rre150m0 ~ duration, data = climate_correlation_data))$r.squared

# summary table of performance (R-squared) 
performance_summary <- tibble(
  Climate_Variable = c("Avg Temp (°C)", "Total Sunshine (Hours)", "Total Precipitation (mm)"),
  R_Squared = c(r_sq_temp, r_sq_sun, r_sq_rain)
) %>%
  # Round R-Squared 
  mutate(R_Squared = round(R_Squared, 3)) %>%
  # Arrange by R-Squared
  arrange(desc(R_Squared)) 

# visualize the R-Squared values
plot_r_squared <- performance_summary %>%
  mutate(Climate_Variable = forcats::fct_reorder(Climate_Variable, R_Squared)) %>%
  ggplot(aes(x = Climate_Variable, y = R_Squared)) +
  geom_col(fill = "#e7298a") +
  coord_flip() +
  geom_text(aes(label = R_Squared), hjust = -0.2, size = 4) +
  labs(
    title = "Predictive Power (R-squared) of Burn Duration by Climate Variable",
    x = NULL,
    y = "$R^2$ Value (Proportion of Variance Explained)"
  ) +
  scale_y_continuous(limits = c(0, max(performance_summary$R_Squared) * 1.2)) +
  theme_minimal(base_size = 14)

# Show the plot
plot_r_squared

# Save 
ggsave(
  filename = here("output", "r_squared_comparison.png"),
  plot = plot_r_squared,
  width = 8, height = 4, dpi = 320, create.dir = TRUE
)
```


